A Causal Model of \hbox{DBMS} Suboptimality\\

### 2017-03-07 subopt-tods16/paper.tex
SubOpt updates

\usepackage{url}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphics}
% \usepackage[total={6.5in,9in}, top=1.25in,left=0.9in]{geometry}
%\usepackage[text={6.5in,9in},centerpage,includefoot]{geometry}
% \usepackage{setspace}
\usepackage{color}
\usepackage{url}
\usepackage{balance}
\usepackage{amsmath}
\usepackage{amssymb}

@@

\section{Introduction}
DBMSes underlie all information systems and hence optimizing their performance is of
critical importance. The DBMS's query optimizer plays an important role. But
what if the optimizer {\em doesn't}: what if it selects the wrong plan?

This paper provides a thorough investigation into \hbox{DBMS} {\em suboptimality},
when the \hbox{DBMS} chooses a slower plan over a faster plan for a given query. We systematically examine the factors influencing the number of suboptimal
queries. There could be multiple causes of the suboptimality. One possible
cause could be some peculiarity within the tens of thousands of lines of
code of that query optimizer. Another possible cause could be the query's
complexity. A third possible reason could be some fundamental limitation
{\em within the general architecture of cost-based optimization} that will
always render a good number of queries suboptimal. Prior research in other domains shows that
increasing the complexity negatively influences
performance~\cite{campbell88,moody98}.

To better understand the impact of different factors on suboptimality of
query performance and the interaction between operators, especially in a
dynamic environment, an experimental approach is needed.
Our research introduces a novel approach
to better understand the factors influencing query performance in
\hbox{DBMSes}. Based on existing research and general knowledge of \hbox{DBMSes}, we
propose an innovative predictive model to understand the presence and influence
of suboptimality in query evaluation. We use an experimental methodology with
the \hbox{DBMS} as a subject to test our hypotheses with respect to factors
influencing suboptimality, applying to 
our experiment data collected over about a \hbox{cumulative} 30~months of query
executions (over 21,000 hours to run almost a million query
executions).\todo{Young: Table I says something different: over 16,000 hours
  ro run almost two million query executions which would correspond to two
  years continuous execution}
Our research falls within creative development of
new evaluation methods and metrics for artifacts, which were identified as
important design-science contributions~\cite{hevner04}.

@@

\item We find that for a surprisingly large portion of queries, the plan
  chosen  by the query optimizer is not the best plan, for some cardinality,
  of the underlying tables.

\item We propose and test a {\em predictive model} for \hbox{DBMSes} to better understand the factors causing suboptimality.

\item \todo{Rick:Mention generational result}

\item The predictive model and these experimental results suggest several specific
  engineering directions.
\end{itemize}

@@

We study these aspects \c2j{}{across \hbox{DBMSes}} to identify the
underlying causes. We have developed an initial predictive causal model that
identifies four 
factors that may play a role in suboptimality. The ultimate goal is to {\em
  understand} a component within a \hbox{DBMS}, its cost-based optimizer, through the
articulation and empirical testing of a general scientific theory.

In Section~\ref{sec:related} we briefly summarize the vast amount of related
work in query optimization to establish the technical basis for our study.
The following section introduces the methodology we will follow, that of
empirical generalization. We then present a predictive, causal model of
suboptimality and empirically test eight specific hypotheses derived from that
model\c2j{. These are the first scientific results that we are aware of that
  test a general causal model of \hbox{DBMSes}.}{ on three \hbox{DBMSes}.
  These are the first scientific results that we are aware of that apply
  {\em across} \hbox{DBMSes}, rather than on a single, specific \hbox{DBMS} or on a
  specific algorithm.} This model has implications for research in
engineering more efficient \hbox{DBMSes}.

@@

We have found through our experiments that flutter and suboptimality are all
around us: {\em every} \hbox{DBMS} that we have examined, including several
open source \hbox{DBMSes} and several proprietary
\hbox{DBMSes}, covering much of the installed base worldwide, exhibit these
phenomena, even when optimizing very simple queries. In the Confirmatory
Experiment described in detail in Section~\ref{sec:experiments}, we started
with 6,966 query instances (a query run on a specific \hbox{DBMS}) after our extensive query measurement
protocol~\doubleblind{\todo{ref SIGMOD'13 paper!}}{}. While about 20\%,
1,491, of these query instances contained only one
plan, a few of the other query instances switched plans at almost every change in cardinality (we varied the
cardinality in increments of 10K tuples, a total of 200 cardinalities): see
Figure~\ref{fig:planchanges}. Slightly over half, 4,093, exhibited
suboptimality somewhere along those 200 cardinalities; a few had
many changes to a plan that was in fact suboptimal, as indicated in
Figure~\ref{fig:suboptplanchanges}, across all four \hbox{DBMSes} considered.

\begin{figure}[h!]\centering
\includegraphics[width=0.60\textwidth]{figures/org_ncpq.pdf}
\caption{Cumulative percentage of queries \hbox{exhibiting} the \hbox{indicated}
  number of plan changes\label{fig:planchanges}}
\end{figure}

\begin{figure}[t]\centering
\includegraphics[width=0.60\textwidth]{figures/zi_ncpq.pdf}
\caption{Cumulative percentage of queries exhibiting the \hbox{indicated}
  number of changes to a {\em sub-optimal}
  plan\label{fig:suboptplanchanges}}
\vspace*{-2ex}
\end{figure}

One oft-stated observation is that the role of query optimization is not to
get the {\em best} plan, but rather to get a plan that is acceptably
good. Figure~\ref{fig:suboptcumulative} shows
the cumulative distribution of the relative amount of suboptimality (where
an $x$-value of 100 denotes that the query ran 100\% slower than the optimal
query, that is, twice as long). The good news is that 2,738 query instances,
or 67\% of the suboptimal queries, exhibited only a small degree of
suboptimality: less than 30\%. The challenge is that over fifth of all
queries (1,355) exhibited a significant amount of suboptimality ($>= 30\%$); the
relative slowdown can be quite large for some queries.

We started with 7,640 query instances (a particular query running on a
particular DBMS, cf.~Experiment 7 of Table~\ref{tab:run_stat}
in Section~\ref{sec:experiments}. After our protocol, we were left with
  6,966 query instances, of which 5,475 had at least one change point, so those are
  the ones we consider further. Of those, 1,382 had {\em no} suboptimality, so
4,093 (75\%) had some suboptimality.

@@

The model also has two interactions within plan space
complexity, between 
CEPS and number of operators with discontinuity variables and between number of repeats and CEPS.

We hypothesize that greater plan space complexity should make it more
difficult to optimize the query.  It is important to note that the
occurrences of sub-optimality is not necessarily dependent on the presence
of a discontinuous plan operator. However, we predict that when
discontinuous plan operators appear in candidate plans, they may introduce
complexity to query optimization, especially at the cardinality where
discontinuity is possible. This is because the performance of such operators
is sensitive to the input size in a rather complex way. Any inaccuracy in
the estimation of plan statistics may lead the optimizer to select a suboptimal
plan.

\vspace{0.6em}\noindent
{\bf Hypothesis 4}: Plan space complexity (CEPS) and Number of operators
with discontinuity will be positively correlated.

@@

\vspace{0.6em}\noindent
{\bf Hypothesis 5}: The Number of repeats will positively correlate with CEPS.

\vspace{0.6em}
The schema complexity construct consists of the variable ``Presence of
secondary indexes.'' Such indexes provide opportunities for the optimizer to
consider more candidate plans that use these indexes, due to the additional
query evaluation operators now applicable. Those additional plans enable the optimizer to
possibly do a better job, while also adding complexity to the optimization
process. We hypothesize that this factor has a more complex role in the
model, serving as a {\em moderator} of two interactions previously
introduced. We hypothesize that the overall effect of the secondary indexes
is to increase the strength of the interaction between query
complexity to plan space complexity.

\vspace{0.6em}\noindent
{\bf Hypothesis 6}: Presence of secondary indexes will strengthen
the correlations between the Query Complexity construct, specifically, the
variables Number of
correlation names, Presence of aggregate, Presence of primary key
attribute, and Presence of subquery, and
the Plan Space Complexity construct, specifically, variables Number of repeats, CEPS, and Number
of operators with discontinuity (correlations a--l listed above in
Hypothesis 2).

@@

\vspace{0.6em}\noindent
{\bf Hypothesis 7}: Suboptimality will be positively correlated with Plan
space complexity, that is, (a)~Number of repeats, (b)~CEPS, and (c)~Number
of operators with discontinuity.

\vspace{0.6em}
We have just {\em described} how
suboptimality might arise, through a theory and its elaborated causal model,
which implies seven specific hypotheses.
We now need to move to {\em prediction}.
How might we test such a model?

@@

In this section we specify how we operationalized each of the seven
variables in the model.  Recall that each independent variable is a property
of a \hbox{DBMS} (number of operators available), of the schema (presence
  of secondary indexes), of a query (number of correlation names,
presence of an aggregate, presence of primary key attribute, and
presence of subquery), of the data (percentage of skewed data), of
the plan space (number of repeats, CEPS, and number of discontinuous
operators). There is 
one dependent variable of our model (suboptimality).

It is important to note that all manipulation must be done {\em
  outside} the \hbox{DBMS}. For proprietary systems we do not have access to the
internal code. We do not know ({\em cannot} know) all the plans that were
considered, nor the details of how the plans are selected. But such access
is not needed; indeed, to be able to study a phenomenon across many \hbox{DBMSes},
such access is not practical. But by designing the experiment to examine
the plans that each \hbox{DBMS} actually produces, and thus to examine phenomena
that can be externally visible, we can obtain valuable insights into general
classes of computational artifacts. Our ultimate goal is to make statements
that hold across cost-based optimizers in general, thereby moving up the
$y$-axis of Figure~\ref{fig:empirical}.

@@

The {\tt EXPLAIN PLAN} facility specifies the operators(s) \hbox{employed} in that
plan. For each \hbox{DBMS}, we collect all the unique operators from all the plans
and count the number of these distinct operators, each used by for least one
query at at least one cardinality by that \hbox{DBMS}. The number of operators used
by a \hbox{DBMS} ranged from 8 to 53 across all the queries and data sets
used in the Exhaustive, Exhaustive with Keys, and Exploratory Experiments,
to be discussed in Section~\ref{sec:experiments}.

@@

%\pagebreak
\shorten{In {\tt generateWhereClause()}, in which all the joins will appear,
  we}We ensure that Cartesian products are
eliminated\c2j{.}{\shorten{({\tt cartesianPossible="false"})}}  We do this by
connecting all the correlation names that appear in the FROM statement via
equi-joins on random attributes. The comparisons are all
equality operators. To ensure that the queries are as simple as possible, we
do not include any additional predicates in the WHERE clause.  \c2j{}{This
  is realized by setting the attributes {\tt maxIsAbsolute} to {\tt true}
  and {\tt complexUsePercentage} to {\tt 100}.  Basically, ``complex''
  predicates eliminate the Cartesian product, and by setting complex
  predicates as ``absolute,'' no additional predicates are included except
  for those which are necessary for eliminating Cartesian product.} Also
for simplicity, we include neither disjunctions nor negations.

@@

\subsection{Data Complexity}\label{sec:datacomplexity}
This independent construct has one variable: percentage of skewed data.
%\todo{Rick:write up operationalization}

@@

We define the skew as the reciprocal of the number of distinct values. 
For 2M distinct values, the skew would be 1/2M, which is practically 0. 
For 1 distinct value, the skew would be 1.0. For 2 distinct values, 
the skew would be 0.5. For 10 distinct values, the skew would be 0.1.

We can generate the table of 2M rows by generating values sequentially from
1 to the number of distinct values.  This creates a "span of values." We
repeat this for the second span if necessary, and on and on, until we have
2M values in all.

@@

What about a skew factor of 0.1? 
This skew factor translates to 10 distinct values. 
So we generate 200K spans, each with 10 values. 
To shorten, we delete 1,000 spans, which is equivalent to removing 10K values. 
As before, this does not change the skew.
At the end, the table will have 1,000 spans with 10K values, retaining a
skew factor of 0.1.

@@

We use five skewness values: 0 (approximately), 0.001, 0.1, 0.5, and 1.0,
for this independent variable.

@@

As discussed in Section~\ref{sec:model}, the ``cardinality of the effective plan
space'' (CEPS) is the number of plans selected as optimal for that query
being evaluated on one of the 200 cardinalities for the variable table. It
is ``effective'' because it was chosen, as opposed to all of the
plans that were considered but not chosen. (Recall that for proprietary
\hbox{DBMSes}, we do not have access to such plans.) Note that we count only
distinct plans. As we saw in Figure~\ref{fig:query769}, fluttering queries
return to a previous plan. This particular query has a CEPS of 4.

@@

To classify each operator as continuous or discontinuous, we use the data
from the Exhaustive Experiment. We
used a small subset of queries to be used later in testing the model, to
attempt to observe the same operators as encountered in that
study. We
collected all query plans at all possible cardinalities (200 in all) and
looked for ``jumps,'' that is, when the cost model for an operator in the
plan exhibited discontinuity. Jumps are determined from the estimated cost
extracted from each plan operator. We provide an example of a jump below. If
a jump is observed, we classify that particular plan operator as
discontinuous.

A {\em jump} is a pair of close cardinalities (in our case, separated by 10K
tuples) in which the query optimizer's cost model is {\em discontinuous},
that is, does not smoothly increase from the lower cardinality to the upper
cardinality. To identify jumps in the first step, we examine the slopes
between each pair of consecutive cardinalities (again, separated by 10K
tuples). We expect that the slope (that is, the first derivative) is
well-behaved and thus does not change much for consecutive cardinalities. 
A jump thus indicates a large deviation in the second derivative of the 
cost model across the two cardinalities.

The Exhaustive experiment gathers the cost model for each operator in each
plan for each cardinality (in our case, for cardinalities ranging from 10K
to 2M in steps of 10K, or 200 points).  Figure~\ref{fig:discontinuity}
presents for a single query\shorten{({\em query4}, exhaustive-10q-1,
  oracle)}, the {\em cost} of each of the hash-join operators utilized in
each plan at a particular cardinality. Each distinct point type depicts an
individual hash-join operator. Note that when two identical query plans
appear at two different cardinalities, we consider all the plan operators
found at the same position within these two plans to be identical. This
figure shows only a small portion of the 200 cardinalities, and is typical
(we generally observed jumps at low cardinalities for these particular queries and
relation sizes).  As shown by this figure, the hash-join operator
represented by `+' has two discontinuous jumps, both appearing below
cardinality 150K. The other hash-join operator represented by $\times$ has one
discontinuous jump.  Between the jumps, the behavior is linear. It is our
guess that
such a jump is caused by the transition between one pass of a disk-based hashing
technique to two passes (and indeed for one of the operators, perhaps the one
lower in the operator tree, the transition to three passes, hence, the two
jumps). That said, all
that matters for our predictive model is that operators that experience
such discontinuity in their cost models present opportunities for
suboptimality.

@@

Formally, we compute the slope ({\em S}) of the {\em estimated} cost
(of the cost model, formalized as $C_{\hbox{\small\em card}}$,
where {\em card} is the input
cardinality) between each pair of consecutive cardinalities as
$S_{\hbox{\small\em card}} = (C_{\hbox{\small\em card} + 10K} -
C_{\hbox{\small\em card}}) / 10K$~.
This computes a series of slope values.
We then compute the standard deviation of all the slope
values.  For example, examining Figure~\ref{fig:discontinuity}, the slopes
are small except at three places, one for the green operator and two for the
red operator. By identifying the slope values that are greater than one
standard deviation over the average value, the discontinuous operators can
be identified: A ``discontinuous operator'' is one for which a jump is
observed in that operator in one of the plans for at least one of the input
queries.

\subsection{Suboptimality}\label{sec:suboptimality}
We now consider the one dependent variable.
How might suboptimality be observed? We have developed a system, \azdb, that
allows us to perform experiments to study this phenomenon of suboptimality.
\hbox{\azdb} submits queries to the \hbox{DBMS}, while varying
the cardinality of one of the tables,
requesting in each case the evaluation plan chosen by the \hbox{DBMS}.
This is done using the
{\tt EXPLAIN} SQL facility available in modern \hbox{DBMSes}. (The Picasso system
also used this facility to visualize the plan space chosen by a \hbox{DBMS}
optimizer~\cite{harish07,Haritsa10}.) We can then compare
the performance (execution time) of various plans for the query, to identify those
situations when a suboptimal plan was chosen, when in fact there was a
different plan that was semantically equivalent to the chosen plan (that is,
yielded the identical result) but which ran faster.

We modify the cardinality to produce multiple execution
plans for a given query. For one of the \hbox{DBMSes}, we can modify the stored
table statistics directly. For the other \hbox{DBMSes}, we had to do so indirectly,
by varying the size of the table and running the optimizer on tables of
different size. As we vary the cardinality, we
collects all the plans that the optimizer felt were appropriate for that
query.

@@

\vspace{1em}
As we will see in Section~\ref{sec:experiments}, we observed only 3,347
violations (0.73\%) of non-strict monotonicity, for the largest experiment
(Experiment 7: Confirmatory) across all
the \hbox{DBMSes} we studied, justifying our conclusion that the \hbox{DBMSes} under
study are indeed monotonic.

@@

No runs violated the experiment-wide sanity checks. For the largest
experiment, the Confirmatory experiment 7 discussed below, approximately
10.5\% of the query exectuions and 5.1\%
of the queryies-at-cardinality (Q@Cs) were dropped due to query execution
and Q@C sanity checks. As a result, excessive variation in calculated query
time was observed in only 0.003\% of the Q@Cs. A total of 0.35\% of Q@C adjacent
pairs violated relaxed monotonicity and 0.68\% of the Q@Cs violated strict
monotonicity, which is acceptable.

>>>>

\usepackage{url}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphics}
% \usepackage[total={6.5in,9in}, top=1.25in,left=0.9in]{geometry}
%\usepackage[text={6.5in,9in},centerpage,includefoot]{geometry}
% \usepackage{setspace}
\usepackage{color}
\usepackage{url}
\usepackage{balance}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{subfigure}

@@

\section{Introduction}\label{sec:intro}
DBMSes underlie information systems and hence optimizing their performance is of
critical importance. The DBMS's query optimizer plays an important role. But
what if the optimizer {\em doesn't}: what if it selects the wrong plan?

This paper provides a thorough investigation into \hbox{DBMS} {\em
  suboptimality}: when the \hbox{DBMS} chooses a slower plan over a faster
plan for a given query. We systematically examine the factors influencing
the number of suboptimal queries. There could be multiple causes of the
suboptimality. One possible cause could be some peculiarity within the tens
of thousands of lines of code of that query optimizer. Another possible
cause could be the query's complexity. Prior research in other domains shows
that increasing complexity negatively influences
performance~\cite{campbell88,moody98}. A third possible reason could be some
fundamental limitation {\em within the general architecture of cost-based
  optimization} that will always render a number of queries suboptimal.

To better understand the impact of different factors on suboptimality of
query performance and the interaction between operators, especially in a
dynamic environment, an experimental approach is needed.
Our research introduces a novel approach
to better understand the factors influencing query performance in
\hbox{DBMSes}. Based on existing research and general knowledge of \hbox{DBMSes}, we
propose an innovative predictive model to better understand the presence and influence
of suboptimality in query evaluation. We use an experimental methodology on
a collection of \hbox{DBMSes} as subjects to test our hypotheses with respect to factors
influencing suboptimality, utilizing
experiment data collected over a \hbox{cumulative} 16,000 hours (over two
years) of query executions.
Our research falls within creative development of
new evaluation methods and metrics for artifacts, which were identified as
important design-science contributions~\cite{hevner04}.

@@

\item We find that for a surprisingly large portion of queries, the plan
  chosen  by the query optimizer is not the best plan, for some cardinality
  of the underlying tables.

\item We propose a {\em predictive model} for \hbox{DBMSes} to better
  understand the factors causing suboptimality.

\item We test the six quite specific hypotheses that arise from that
  model across a wide range of DBMSes, queries, and data.

\item We show through correlational and regression analysis that four
  hypothesized interactions are strongly supported, one is partially supported, and
  an interaction hypothesized to go one direction went quite weakly the
  other direction, lending credibility to our particular model.

\item Through an innovative
  analysis, we track the net cumulative benefit of a succession of operators
  added to the DBMS, and show that current relational DBMSes may have
  already reached the stage where it is difficult to provide performance
  improvement via new operators, and in fact performance may degrade.

\item The predictive model and these experimental results suggest several specific
  engineering directions.

\item The model and analyses presented here taken as a whole imply that a
  new approach to query optimization, fundamentally
  different from that utilized over the last forty years, may be required
  to get past a fundamental limitation uncovered in this research.
\end{itemize}

@@

We study these aspects \c2j{}{across \hbox{DBMSes}} to identify the
underlying causes. We have developed a predictive causal model that
identifies four 
factors that may play a role in suboptimality. The ultimate goal is to {\em
  understand} a component within a \hbox{DBMS}, its cost-based optimizer, through the
articulation and empirical testing of a general scientific theory.

In Section~\ref{sec:related} we briefly summarize the vast amount of related
work in query optimization to establish the technical basis for our study.
The following section introduces the methodology we will follow, that of
{\em empirical generalization}~\cite{cohenbook}. We present in
Sections~\ref{sec:model} and~\ref{sec:operationalization} a predictive,
causal model of suboptimality and state seven specific hypotheses derived
from that model. We then test these hypotheses across a number of queries,
data, and cardinalities, whose results provide strong support for the
validity of the proposed model.  These are the first scientific results that
we are aware of that apply {\em across} \hbox{DBMSes}, rather than on a
single, specific \hbox{DBMS} or on a specific algorithm. In
Section~\ref{sec:diminishing} presents a follow-on analysis that
demonstrates a fundamental limit to the number of operators that a
cost-based optimizer can support, and indicates that perhaps extant DBMSes
have hit that limit. This model has implications listed in
Section~\ref{sec:engineering} for research in engineering more efficient
\hbox{DBMSes}.

@@

We have found through our experiments that flutter and suboptimality are all
around us: {\em every} \hbox{DBMS} that we have examined, including several
open source \hbox{DBMSes} and several proprietary
\hbox{DBMSes}, covering much of the installed base worldwide, exhibit these
phenomena, even when optimizing very simple queries. In the Confirmatory
Experiment described in detail in Section~\ref{sec:experiments}, we started
with 6,967 query instances (a query run on a specific \hbox{DBMS}) after \doubleblind{our}{an} extensive query measurement
protocol~\cite{TTPv1,TTPv2} applied its extensive sanity checks. While about
20\%, or 1,491, of these query instances contained only one
plan, a few of the other query instances switched plans at almost every change in cardinality (we varied the
cardinality in increments of 10K tuples, a total of 200 cardinalities): see
Figure~\ref{fig:planchanges}. Slightly over half, or 3,933 query instances, exhibited
suboptimality somewhere along those 200 cardinalities; a few had
many changes to a plan that was in fact suboptimal, as indicated in
Figure~\ref{fig:suboptplanchanges}, across all four \hbox{DBMSes} considered.

\begin{figure}[h!]\centering
%\todo{Young: were is the figure!}
\includegraphics[width=0.60\textwidth]{figures/org_ncpq.pdf}
\caption{Cumulative percentage of queries \hbox{exhibiting} the \hbox{indicated}
  number of plan changes\label{fig:planchanges}}
\end{figure}

\begin{figure}[t]\centering
\includegraphics[width=0.60\textwidth]{figures/zi_ncpq.pdf}
\caption{Cumulative percentage of queries exhibiting the \hbox{indicated}
  number of changes to a {\em suboptimal} plan\label{fig:suboptplanchanges}}
\vspace*{-2ex}
\end{figure}

One oft-stated observation is that the role of query optimization is not to
get the {\em best} plan, but rather to get a plan that is acceptably
good. (Thus, the very term ``query optimizer'' is aspirational rather than
accurate.) Figure~\ref{fig:suboptcumulative} shows
the cumulative distribution of the relative amount of suboptimality (where
an $x$-value of 100 denotes that the query ran 100\% slower than the optimal
query, that is, twice as long). The good news is that 2,738 query instances,
or 67\% of the suboptimal queries, exhibited only a small degree of
suboptimality: less than 30\%. The challenge is that over fifth of all
queries (1,355) exhibited a significant amount of suboptimality ($\geq
30\%$). The relative slowdown can thus be quite large for some queries.

We started with 7,640 query instances, each of which is a particular query running on a
particular DBMS, cf.~Experiment 7 of Table~\ref{tab:run_stat}
in Section~\ref{sec:experiments}. After our protocol, we were left with
  6,967 query instances, of which 5,475 had at least one change point, so those are
  the ones we consider further. Of those, 1,382 had {\em no} suboptimality, so
4,093 (75\%) had some suboptimality.

@@

The model also has two interactions within plan space
complexity, between 
CEPS and number of operators with discontinuity variables and between number of repeats.

We hypothesize that greater plan space complexity should make it more
difficult to optimize the query.  It is important to note that the
occurrences of suboptimality is not necessarily dependent on the presence
of a discontinuous plan operator. However, we predict that when
discontinuous plan operators appear in candidate plans, they may introduce
complexity to query optimization, especially at the cardinality where
discontinuity is possible. This is because the performance of such operators
is sensitive to the input size in a rather complex way. Any inaccuracy in
the estimation of plan statistics may lead the optimizer to select a suboptimal
plan.

@@

\vspace{0.6em}\noindent
{\bf Hypothesis 4}: Plan space complexity (CEPS) and (a)~Number of operators
with discontinuity and (b) Number of repeats will be positively correlated.

\vspace{0.6em}
The schema complexity construct consists of the variable ``Presence of
secondary indexes.'' Such indexes provide opportunities for the optimizer to
consider more candidate plans that use these indexes, due to the additional
query evaluation operators now applicable. Those additional plans enable the optimizer to
possibly do a better job, while also adding complexity to the optimization
process. We hypothesize that this factor has a more complex role in the
model, serving as a {\em moderator} of two interactions previously
introduced. We hypothesize that the overall effect of the secondary indexes
is to increase the strength of the interaction between query
complexity to plan space complexity. (Note that as the presence of primary
key is required for secondary indexes, we don't include that former
independent variable here.)

\vspace{0.6em}\noindent
{\bf Hypothesis 5}: Presence of secondary indexes will strengthen
the correlations between the Query Complexity construct, specifically, the
variables Number of
correlation names, Presence of aggregate, and Presence of subquery, and
the Plan Space Complexity construct, specifically, variables Number of repeats, CEPS, and Number
of operators with discontinuity (correlations a--f and j--l listed above in
Hypothesis 2).

@@

\vspace{0.6em}\noindent
{\bf Hypothesis 6}: Suboptimality will be positively correlated with Plan
space complexity, that is, (a)~Number of repeats, (b)~CEPS, and (c)~Number
of operators with discontinuity.

\vspace{0.6em}
We have just {\em described} how
suboptimality might arise, through a theory and its elaborated causal model,
which implies six specific hypotheses (some with multiple components).
We now need to move to {\em prediction}.
How might we test such a model?

@@

In this section we specify how we operationalized each of the seven
variables in the model.  Recall that each independent variable is a property
of a \hbox{DBMS} (number of operators available), of the schema (presence
  of secondary indexes), of a query (number of correlation names,
presence of an aggregate, presence of primary key attribute, and
presence of subquery), of the data (percentage of skewed data), and of
the plan space (number of repeats, CEPS, and number of discontinuous
operators). There is 
one dependent variable of our model (suboptimality).

It is important to note that manipulation must be done {\em
  outside} the \hbox{DBMS}, as we do not have access to the
internal code of proprietary DBMSes. Hence, we do not know ({\em cannot} know) all the plans that were
considered, nor the details of how the plans are selected. But such access
is not needed; indeed, to be able to study a phenomenon across many \hbox{DBMSes},
such access is not practical. But by designing the experiment to examine
the plans that each \hbox{DBMS} actually produces, and thus to examine phenomena
that can be externally visible, we can obtain valuable insights into general
classes of computational artifacts. Our ultimate goal is to make statements
that hold across cost-based optimizers in general, thereby moving up the
$y$-axis of Figure~\ref{fig:empirical}.

@@

The {\tt EXPLAIN PLAN} facility specifies the operators(s) \hbox{employed} in that
plan. For each \hbox{DBMS}, we collect the unique operators from the plans
and count the number of these distinct operators, each used by for least one
query at at least one cardinality by that \hbox{DBMS}. The number of operators used
by a \hbox{DBMS} ranged from 8 to 53 across the queries and data sets
used in the Exhaustive, Exhaustive with Keys, and Exploratory Experiments,
to be discussed in Section~\ref{sec:experiments}.

@@

%\pagebreak
\shorten{In {\tt generateWhereClause()}, in which all the joins will appear,
  we}We ensure that Cartesian products are
eliminated\c2j{.}{\shorten{({\tt cartesianPossible="false"})}}  We do this by
connecting the correlation names that appear in the FROM statement via
equi-joins on random attributes. The comparisons are all
equality operators. To ensure that the queries are as simple as possible, we
do not include any additional predicates in the WHERE clause.  \c2j{}{This
  is realized by setting the attributes {\tt maxIsAbsolute} to {\tt true}
  and {\tt complexUsePercentage} to {\tt 100}.  Basically, ``complex''
  predicates eliminate the Cartesian product, and by setting complex
  predicates as ``absolute,'' no additional predicates are included except
  for those which are necessary for eliminating Cartesian product.} Also
for simplicity, we include neither disjunctions nor negations.

@@

\subsection{Data Complexity}\label{sec:datacomplexity}
This independent construct has one variable: percentage of skewed data.

@@

We define the skew quite simply: ``the reciprocal of the number of distinct
values,'' so $0 < \hbox{\em skew} \leq 1$.
For 2M distinct values, the skew would be $\frac{1}{2M}= 0.000005$, which is practically 0. 
For exactly one distinct value, the skew would be 1.0. For two distinct values, 
the skew would be 0.5. For ten distinct values, the skew would be 0.1.

We can generate the table of 2M rows by generating values sequentially from
1 to the number of distinct values.  This creates a ``span of values.'' We
repeat this for the second span if necessary, and on and on, until we have
2M values in all.

@@

What about a skew factor of 0.1? 
This skew factor translates to 10 distinct values. 
So we generate 200K spans, each with ten values. 
To shorten, we delete 1,000 spans, which is equivalent to removing 10K values. 
As before, this does not change the skew.
At the end, the table will have 1,000 spans with 10K values, retaining a
skew factor of 0.1.

@@

We use five skewness values: 0.000005 (which we term only slightly
misleadingly as ``no skew''), 0.001, 0.1, 0.5, and 1.0,
for this independent variable.

@@

As discussed in Section~\ref{sec:model}, the ``cardinality of the effective plan
space'' (CEPS) is the number of plans selected as optimal for that query
being evaluated on one of the 200 cardinalities for the variable table. It
is ``effective'' because it was chosen, as opposed to of the
plans that were considered but not chosen. (Recall that for proprietary
\hbox{DBMSes}, we do not have access to such plans.) Note that we count only
distinct plans. As we saw in Figure~\ref{fig:query769}, fluttering queries
return to a previous plan. This particular query has a CEPS of 4.

@@

To classify each operator as continuous or discontinuous, we use the data
from the Exhaustive Experiment. We
used a small subset of queries to be used later in testing the model, to
attempt to observe the same operators as encountered in that
study. We
collected the query plan at each possible cardinalities (200 in all) and
looked for ``jumps,'' that is, when the cost model for an operator in the
plan exhibited discontinuity. Jumps are determined from the estimated cost
extracted from each plan operator. We provide an example of a jump below. If
a jump is observed, we classify that particular plan operator as
discontinuous.

A {\em jump} is a pair of close cardinalities (in our case, separated by 10K
tuples) in which the query optimizer's cost model is {\em discontinuous},
that is, does not smoothly increase from the lower cardinality to the upper
cardinality. To identify jumps in the first step, we examine the slopes
between each pair of adjacent cardinalities (again, separated by 10K
tuples). We expect that the slope (that is, the first derivative) is
well-behaved and thus does not change much for adjacent cardinalities. 
A jump thus indicates a large deviation in the second derivative of the 
cost model across the two cardinalities.

The Exhaustive experiment gathers the cost model for each operator in each
plan for each cardinality (in our case, for cardinalities ranging from 10K
to 2M in steps of 10K, or 200 points).  Figure~\ref{fig:discontinuity}
presents for a single query\shorten{({\em query4}, exhaustive-10q-1,
  oracle)}, the {\em cost} of each of the hash-join operators utilized in
each plan at a particular cardinality. Each distinct point type depicts an
individual hash-join operator. \todo{Young: I don't understand what this
  sentence adds'' Rick: Yes, please delete the sentence.}Note that when two identical query plans
appear at two different cardinalities, we consider all the plan operators
found at the same position within these two plans to be identical. This
figure shows only a small portion of the 200 cardinalities, and is typical
(we generally observed jumps at low cardinalities for these particular queries and
relation sizes).  As shown by this figure, the hash-join operator
represented by `+' has two discontinuous jumps, both appearing below
cardinality 150K. The other hash-join operator represented by $\times$ has one
discontinuous jump.  Between the jumps, the behavior is linear. It is our
guess that
such a jump is caused by the transition between one pass of a disk-based hashing
technique to two passes (and indeed for one of the operators, perhaps the one
lower in the operator tree, the transition to three passes, hence, the two
jumps). That said, all
that matters for our predictive model is that operators that experience
such discontinuity in their cost models present opportunities for
suboptimality.

@@

Formally, we compute the slope ({\em S}) of the {\em estimated} cost
(of the cost model, formalized as $C_{\hbox{\small\em card}}$,
where {\em card} is the input
cardinality) between each pair of adjacent cardinalities as
$S_{\hbox{\small\em card}} = (C_{\hbox{\small\em card} + 10K} -
C_{\hbox{\small\em card}}) / 10K$~.
This computes a series of slope values.
We then compute the standard deviation of the slope
values.  For example, examining Figure~\ref{fig:discontinuity}, the slopes
are small except at three places, one for the green operator and two for the
red operator. By identifying the slope values that are greater than one
standard deviation over the average value, the discontinuous operators can
be identified: A ``discontinuous operator'' is one for which a jump is
observed in that operator in one of the plans for at least one of the input
queries.

\subsection{Suboptimality}\label{sec:suboptimality}
We now consider the one dependent variable at the core of this investigation.
How might suboptimality be observed? We have developed a system, \azdb, that
allows us to perform experiments to study this phenomenon of suboptimality.
\hbox{\azdb} submits queries to the \hbox{DBMS}, while varying
the cardinality of one of the tables,
requesting in each case the evaluation plan chosen by the \hbox{DBMS}.
This is done using the
{\tt EXPLAIN} SQL facility available in modern \hbox{DBMSes}. (The Picasso system
also used this facility to visualize the plan space chosen by a \hbox{DBMS}
optimizer~\cite{harish07,Haritsa10}.) We can then compare
the performance (execution time) of various plans for the query, to identify those
situations when a suboptimal plan was chosen, when in fact there was a
different plan that was semantically equivalent to the chosen plan (that is,
yielded the identical result) but which ran faster.

We modify the cardinality to produce multiple execution plans for a given
query. For one of the \hbox{DBMSes}, we can modify the stored table
statistics directly. For the other \hbox{DBMSes}, we had to do so
indirectly, by varying the size of the table and running the optimizer on
tables of different size. As we vary the cardinality, we collect the plans
that the optimizer felt were appropriate for that query at the various cardinalities.

@@

\vspace{1em}
As we will see in Section~\ref{sec:experiments}, we observed only 3,347
violations (0.74\%) of non-strict monotonicity, for the largest experiment
(Experiment 7: Confirmatory) across all
the \hbox{DBMSes} we studied, justifying our conclusion that the \hbox{DBMSes} under
study are indeed monotonic.

@@

No run violated the experiment-wide sanity checks. For the largest
experiment, the Confirmatory experiment 7 discussed below, approximately
10.5\% of the query executions and 5.1\%
of the queries-at-cardinality (Q@Cs) were dropped due to query execution
and Q@C sanity checks. As a result, excessive variation in calculated query
time was observed in only 0.003\% of the Q@Cs. A total of 0.43\% of Q@C adjacent
pairs violated relaxed monotonicity and 0.74\% of the Q@Cs violated strict
monotonicity, which is acceptable.
